Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-30 paper.tex
Update saige to 10_5

In this article, we decouple the VCF data model from its row-oriented
file definition, and show how the data can be
compactly stored and efficiently analysed in a cloud-native, FAIR manner.
We do this by translating VCF data into Zarr format,
a method of storing large-scale multidimensional data as a regular
grid of compressed chunks.
Zarr's elegant simplicity and first-class support for
cloud object stores and parallel computing have led to
it gaining substantial traction
across the sciences, and it is now used in multiple petabyte-scale
datasets in cloud deployments (see Methods for details).
We present the VCF Zarr specification that formalises this
mapping, along with the \texttt{vcf2zarr} and \texttt{vcztools} 
utilities to reliably convert VCF to and from Zarr at scale.
We show that VCF Zarr is much more compact than
VCF and is competitive with state-of-the-art
file-based VCF compression tools.
Moreover, we show that Zarr's storage of data in an analysis-ready
format greatly facilitates computation,
with various benchmarks being substantially faster than
\texttt{bcftools} based pipelines, and again competitive
with state-of-the-art file-oriented methods. 
We demonstrate the scalability and flexibility of VCF Zarr
via case-studies on five boundary-pushing datasets.
We examine conversion and compression performance on three different
data modalities for large human datasets: whole genome sequence
data from Genomics England (5X smaller than gzipped VCF);
genotype data from Our Future Health (2.5X smaller than gzipped VCF);
and exome-like data from All of Us (1.1X larger than gzipped VCF). 
We then examine performance 
for species with large genomes using Norway Spruce data (3.2X smaller
than gzipped VCF),
and in a large collection of short whole genome alignments
using SARS-CoV-2 data 
(81X smaller than gzipped FASTA). 
Interoperability with the vast ecosystem of tools
based on VCF~\citep{danecek2021twelve,garrison2022spectrum}
is crucial. 
% TODO sentence too long
We first present the \texttt{vcztools} program,
which implements a subset of the functionality in \texttt{bcftools}
and allows data stored in Zarr at some URL to be 
treated as if it were a VCF or BCF file,
and demonstrate that it can be used as a practical replacement 
in processing pipelines.
We then present an illustrative example of how existing
methods can adopt VCF Zarr by modifying the popular 
SAIGE~\cite{zhou2018efficiently, zhou2020scalable} GWAS software
(written in C++), and show significant speedups over the existing 
VCF (6.5X) and BCF (3X) backends in a simple benchmark.
Finally, we illustrate the potential 
for a new generation of Zarr-based applications via some illustative
benchmarks. First, we demonstrate the very high levels of throughput 
performance available on public clouds when data in an object 
store is accessed in an asynchronous and highly parallel manner.
Finally, we demonstrate the ease and efficiency with which data
in Zarr format can be processed on a GPU, opening many exciting
avenues for future research.

@@

\subsection{Interoperability: SAIGE proof-of-concept}
To demonstrate the potential for integrating VCF Zarr into existing
high-performance analysis tools,
we developed a proof-of-concept extension of SAIGE~\cite{zhou2018efficiently,
zhou2020scalable}, a popular package for performing genome wide
association tests. 
Using the TensorStore C++ library~\cite{tensorstore}
we added support for VCF Zarr via an additional backend 
(SAIGE reading data from PLINK, BGEN, VCF, BCF and SAV formats) 
for which the core functionality is defined in a 264 line C++ file.

To assess the performance of this implementation we compared the time required
to perform an association test using the simulated dataset of 10,000 samples
from Fig~\ref{fig-data-storage}. We performed a single-variant association
test based on a phenotype simulated using
\texttt{tstrait}~\citep{tagami2024tstrait},
and measure the overall time required when using the 
equivalent VCF, BCF, SAV and Zarr files.

We compared the same analysis using the existing VCF, BCF and Savvy backends
to our prototype Zarr backend.
[Times: BCF 3m, VCF6.5m, Savvy 39s, Zarr 55s]

% BCF:
% real 3m1.068s user 2m56.944s sys 0m2.941s

% VCF:
% real 6m30.335s user 6m25.193s sys 0m2.842s

% Savvy:
% real 0m39.660s user 0m37.498s sys 0m2.183s

% Zarr:
% real 0m55.957s user 1m10.375s sys 0m12.111s

>>>>

In this article, we decouple the VCF data model from its row-oriented
file definition, and show how the data can be
compactly stored and efficiently analysed in a cloud-native, FAIR manner.
We do this by translating VCF data into Zarr format,
a method of storing large-scale multidimensional data as a regular
grid of compressed chunks.
Zarr's elegant simplicity and first-class support for
cloud object stores and parallel computing have led to
it gaining substantial traction
across the sciences, and it is now used in multiple petabyte-scale
datasets in cloud deployments (see Methods for details).
We present the VCF Zarr specification that formalises this
mapping, along with the \texttt{vcf2zarr} and \texttt{vcztools} 
utilities to reliably convert VCF to and from Zarr at scale.
We show that VCF Zarr is much more compact than
VCF and is competitive with state-of-the-art
file-based VCF compression tools.
Moreover, we show that Zarr's storage of data in an analysis-ready
format greatly facilitates computation,
with various benchmarks being substantially faster than
\texttt{bcftools} based pipelines, and again competitive
with state-of-the-art file-oriented methods. 
We demonstrate the scalability and flexibility of VCF Zarr
via case-studies on five boundary-pushing datasets.
We examine conversion and compression performance on three different
data modalities for large human datasets: whole genome sequence
data from Genomics England (5X smaller than gzipped VCF);
genotype data from Our Future Health (2.5X smaller than gzipped VCF);
and exome-like data from All of Us (1.1X larger than gzipped VCF). 
We then examine performance 
for species with large genomes using Norway Spruce data (3.2X smaller
than gzipped VCF),
and in a large collection of short whole genome alignments
using SARS-CoV-2 data 
(81X smaller than gzipped FASTA). 
Interoperability with the vast ecosystem of tools
based on VCF~\citep{danecek2021twelve,garrison2022spectrum}
is crucial. 
% TODO sentence too long
We first present the \texttt{vcztools} program,
which implements a subset of the functionality in \texttt{bcftools}
and allows data stored in Zarr at some URL to be 
treated as if it were a VCF or BCF file,
and demonstrate that it can be used as a practical replacement 
in processing pipelines.
We then present an illustrative example of how existing
methods can adopt VCF Zarr by modifying the popular 
SAIGE~\cite{zhou2018efficiently, zhou2020scalable} GWAS software
(written in C++), and show significant speedups over the existing 
VCF (7.9X) and BCF (3.4X) backends in a simple benchmark.
Finally, we illustrate the potential 
for a new generation of Zarr-based applications via some illustative
benchmarks. First, we demonstrate the very high levels of throughput 
performance available on public clouds when data in an object 
store is accessed in an asynchronous and highly parallel manner.
Finally, we demonstrate the ease and efficiency with which data
in Zarr format can be processed on a GPU, opening many exciting
avenues for future research.

@@

\subsection{Interoperability: SAIGE proof-of-concept}
To demonstrate the potential for integrating VCF Zarr into existing
high-performance analysis tools,
we developed a proof-of-concept extension of SAIGE~\cite{zhou2018efficiently,
zhou2020scalable}, a popular package for performing genome wide
association tests. 
Using the TensorStore C++ library~\cite{tensorstore}
we added support for VCF Zarr via an additional backend 
(SAIGE reads data from PLINK, BGEN, VCF, BCF and SAV formats) 
for which the core functionality is defined in a 264 line C++ file.

To assess the performance of this implementation we compared the time required
to perform an association test using the simulated dataset of 100,000 samples
from Fig~\ref{fig-data-storage}. We performed a single-variant association
test based on a phenotype simulated using
\texttt{tstrait}~\citep{tagami2024tstrait},
and measure the overall time required when using the 
equivalent VCF, BCF, SAV and Zarr files.

We compared the same analysis using the existing VCF, BCF and Savvy backends
to our prototype Zarr backend.
[Times: BCF 70m, VCF 162m, Savvy 7m, Zarr 20m]

%BCF:
%user   system  elapsed 
%4195.218    4.671 4200.508
%VCF:
%user  system elapsed 
%9718.428    4.735 9727.995 
%Savvy:
%user  system elapsed 
%415.371   2.420 418.098 
%Zarr:
%user   system  elapsed 
%1758.976  259.924 1223.767
